---
title: 'EDA Project: MLB data'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This project will begin on Monday June 7th, and __conclude with a 10-15 minute presentation one week later on Thursday, June 17th during lab from 2:30 to 4 PM EDT__. The goal of this project is to practice understanding the structure of a dataset, and to practice generating and evaluating hypotheses using fundamental EDA and data visualization techniques.

## Deliverables

Your team is expected to produce `R Markdown` slides (an example template will be provided shortly) to accompany your 10-15 minute presentation with the following information:

* Explanation of the data structure of the dataset,
* __Three hypotheses__ you are interested in exploring,
* __Three data visualizations__ exploring the hypotheses, at least two of which must be multivariate.  __Each visualization must be in a different format__ from the other two, and you must have at least one categorical and one continuous visualization.
* __One clustering example__,
* Conclusions reached for the hypotheses based on your EDA and data visualizations,

## Timeline

There will be two submission deadlines:

**Friday, June 11th @ 4:00pm EST** - Each student will push their individual code for the project thus far to their GitHub accounts for review. We will then provide feedback on the code submitted.

**Thursday, June 17 @ 12:00pm EST** - Slides and full code must be completed and ready for presentation. Send your slides to Ron's email ([ryurko@andrew.cmu.edu](mailto:ryurko@andrew.cmu.edu)).  All code, visualizations, and presentations must be made in `R`. Take advantage of examples from lecture and the presentation template, but also feel free to explore material online that may be relevant!


## Data

Your team is assigned the [__MLB batted balls data__](http://www.stat.cmu.edu/cmsac/sure/2021/materials/data/eda_projects/mlb_batted_balls_2021.csv). This dataset contains all batted balls from the current 2021 MLB season, through June 5th, courtesy of [baseballsavant.com](https://baseballsavant.mlb.com/) and accessed using the [`baseballr`](http://billpetti.github.io/baseballr/) package. The code chunk at the end shows how this dataset was constructed in `R`.

Each row of the dataset corresponds to a batted ball and has the following columns:

* `player_name`: Name of the batter in Last, First format
* `batter`: Unique identifier for the batter
* `stand`: handedness of hitter, either `L` (left) or `R` (right), note that switch hitters are in this table but they'll switch depending on the pitcher,
* `events`: categorical event denoting the outcome of the batted ball,
* `hc_x`: horizontal coordinate of where a batted ball is first touched by a fielder,
* `hc_y`: vertical coordinate of where a batted ball is first touched by a fielder (note you should take the multiply this by -1 when plotting),
* `hit_distance_sc`: distance of the batted ball in feet according to Statcast,
* `launch_speed`: exit velocity of the ball off the bat (mph),
* `launch_angle`: the vertical angle of the ball off the bat measured from a line parallel to the ground,
* `hit_location`: positional number of the player who fielded the ball, possible values are 1-9 for each player (see [here](https://en.wikipedia.org/wiki/Baseball_positions)),
* `bb_type`: batted ball type,
* `barrel`: indicator if the batted ball was a ["barrel"](https://www.mlb.com/glossary/statcast/barrel),
* `pitch_type`: type of pitch thrown according to MLB's algorithm,
* `release_speed`: speed of the pitch measured when the ball is released (mph),
* `effective_speed`: perceived velocity of the ball, i.e., the velocity of the pitch is adjusted for how close it is to home when it is released (mph)
* `if_fielding_alignment`: type of infield shift by defense
* `of_fielding_alignment`: type of outfield shift by defense
* `game_date`: date of the game (mm/dd/yyyy)
* `balls`: number of balls in the count,
* `strikes`: number of strikes in the count,
* `outs_when_up`: number of outs when the batter is up,
* `on_1b`: unique identifier for a runner on first base (if there is one),
* `on_2b`: unique identifier for a runner on second base (if there is one),
* `on_3b`: unique identifier for a runner on third base (if there is one),
* `inning`: the inning number,
* `inning_topbot`: top or bottom of the inning,
* `home_score`: home team score before batted ball,
* `away_score`: away team score before batted ball,  
* `post_home_score`: home team score after batted ball,
* `post_away_score`: away team score after batted ball,  
* `des`: description of the batted ball and play.

Note that a full glossary of the features available from MLB's Statcast data can be found [here](https://app.box.com/v/statcast-pitchfx-glossary-pett).


## Code to build dataset

```{r}
# install.packages("devtools")
# devtools::install_github("BillPetti/baseballr")
```



```{r}
library(baseballr)
library(tidyverse)

s1 <- scrape_statcast_savant_batter_all("2021-04-01", 

                                        "2021-04-07") %>%
  dplyr::filter(type == "X")

s2 <- scrape_statcast_savant_batter_all("2021-04-08", 

                                        "2021-04-14") %>%
  dplyr::filter(type == "X")

s3 <- scrape_statcast_savant_batter_all("2021-04-15", 

                                        "2021-04-22") %>%
  dplyr::filter(type == "X")

s4 <- scrape_statcast_savant_batter_all("2021-04-23", 

                                        "2021-04-30") %>%
  dplyr::filter(type == "X")


s5 <- scrape_statcast_savant_batter_all("2021-05-01",

                                        "2021-05-06") %>%
  dplyr::filter(type == "X")

s6 <- scrape_statcast_savant_batter_all("2021-05-07",

                                        "2021-05-13") %>%
  dplyr::filter(type == "X")

s7 <- scrape_statcast_savant_batter_all("2021-05-14",

                                        "2021-05-20") %>%
  dplyr::filter(type == "X")

s8 <- scrape_statcast_savant_batter_all("2021-05-21",

                                        "2021-05-27") %>%
  dplyr::filter(type == "X")

s9 <- scrape_statcast_savant_batter_all("2021-05-28",

                                        "2021-06-03") %>%
  dplyr::filter(type == "X")

s10 <- scrape_statcast_savant_batter_all("2021-06-04",

                                         "2021-06-10") %>%
  dplyr::filter(type == "X")

s11 <- scrape_statcast_savant_batter_all("2021-06-11",

                                         "2021-06-17") %>%
  dplyr::filter(type == "X")

mlb_batted_balls_2021 <- rbind(s1, s2, s3, s4, s5, s6, s7, s8, s9, s10,s11)
```

```{r}
unique(mlb_batted_balls_2021$game_date)
```

```{r, eval = FALSE}
library(baseballr)
library(tidyverse)

# Scrape all data for this season:
# mlb_batted_balls_2021 <-
#   scrape_statcast_savant_batter_all(start_date = "2021-05-31",
#                                     end_date = "2021-06-20") %>%
#   dplyr::filter(type == "X")

mlb_batted_balls_2021 <- mlb_batted_balls_2021 %>%
  # Only select columns regarding the batted ball with discrete pitch type
  # information (except for the speed) for now: 
  dplyr::select(# Batter info:
                player_name, batter, stand, 
                # Batted ball info:
                events, hc_x, hc_y, hit_distance_sc, launch_speed, launch_angle,
                hit_location, bb_type, barrel,
                # Pitch info:
                pitch_type, release_speed, effective_speed, 
                # Shift info:
                if_fielding_alignment, of_fielding_alignment,
                # Game level context:
                game_date, balls, strikes, outs_when_up, on_1b, on_2b, on_3b, 
                inning, inning_topbot, home_score, away_score, post_home_score,
                post_away_score, home_team, away_team,
                # Description of play:
                des)

# if inning_topbot is Top that row corresponds to the away team
# if inning_topbot is Bot that row corresponds to the home team
mlb_batted_balls_2021

```



```{r}
mlb_batted_balls_2021$row_team <- 
  ifelse(mlb_batted_balls_2021$inning_topbot == 'Top', mlb_batted_balls_2021$away_team, 
         mlb_batted_balls_2021$home_team)
head(mlb_batted_balls_2021)
```

```{r}
unique(mlb_batted_balls_2021$game_date)
```


```{r}
length(unique(mlb_batted_balls_2021$row_team))
```


```{r}
#devtools::install_github("bdilday/GeomMLBStadiums")
library(GeomMLBStadiums)
library(ggplot2)
library(dplyr)
```

```{r}
head(MLBStadiumsPathData)
```


```{r}
length(unique(MLBStadiumsPathData$team))
```

```{r}
unique(MLBStadiumsPathData$team)
```

```{r}
ggplot() + 
  geom_mlb_stadium(stadium_ids = "all_mlb", 
                   stadium_segments = "all") + 
  facet_wrap(~team) + 
  coord_fixed() + 
  theme_void()
```


```{r}
# my_list <- list('OAK' = 'angels',
#                 name2 = your_comp2)
```

```{r}
unique(mlb_batted_balls_2021$row_team)
```

```{r}
keys <- c("OAK", "LAD", "KC",  "MIA", "STL", "COL", "ARI", "SEA", "NYY", "MIN", "TB",  
          "SD",  "TEX", "BOS" ,"CWS" ,"SF"  ,"ATL", "TOR", "MIL" ,"PHI", "PIT", "DET", 
          "CHC", "NYM", "CIN" ,"BAL", "WSH", "CLE", "HOU","LAA")
values <- c("athletics","dodgers", "royals","marlins","cardinals","rockies","diamondbacks","mariners","yankees","twins","rays","padres","rangers","red sox",
            "white sox","giants","braves","blue jays","brewers","phillies","pirates","tigers","cubs","mets","reds","orioles","nationals","indians","astros",
            "angels")
names(values) <- keys
my_list <- as.list(values)
my_list
```



```{r}
subset <- unique(mlb_batted_balls_2021[c("home_team", "away_team")])
subset
```


```{r}
subset2 <- subset[!duplicated(t(apply(subset, 1, sort))),]
```


```{r}
# rowwise <- function(dat) unique(t(apply(dat, 1, sort)))
# subset2 <- data.frame(rowwise(subset))
# names(subset2)[1] <- "home_team"
# names(subset2)[2] <- "away_team"
# subset2
```

```{r}
library(dplyr)
x1 <- setdiff(subset, subset2)
x1
```

```{r}
library(data.table)
x2 <- copy(x1)
x2[1:2] <- x2[2:1]
x2
```

```{r}
valid_games <- rbind(x1, x2)
valid_games$identifier <- with(valid_games, paste0(home_team, away_team))
valid_games
```

```{r}
mlb_batted_balls_2021$identifier <- with(mlb_batted_balls_2021, paste0(home_team, away_team))
```


```{r}
valid_data <- subset(mlb_batted_balls_2021, identifier %in% valid_games$identifier)
```

```{r}
valid_data <- read.csv('valid_data.csv')
```

```{r}
dim(valid_data)
```


```{r}
unique(valid_data[c("home_team", "away_team")])
```

```{r}
valid_data$month <- months(as.Date(valid_data$game_date))
unique(valid_data$month)
```

```{r}
write.csv(mlb_batted_balls_2021, "C:\\Users\\avija\\Documents\\CMSAC_2021\\cmsacamp-eda-project\\mlb_data.csv", 
          row.names = FALSE)
```

```{r}
head(mlb_batted_balls_2021)
```


```{r}
# this gives us baseball data that was PLAYED in COL Field
test_data <- mlb_batted_balls_2021 %>% filter(home_team == 'COL')
batted_ball_data <- copy(test_data)
unique(batted_ball_data$game_date)
dates <- length(unique(batted_ball_data$game_date))
```

```{r}
batted_ball_data$team <- rep(c(my_list[[batted_ball_data$home_team[1]]]), each=dim(batted_ball_data)[1])

# batted_ball_data$row_team <- factor(batted_ball_data$row_team, levels = c("COL", "ARI"))

batted_ball_data %>% mlbam_xy_transformation() %>% 
    ggplot(aes(x = hc_x_, y = hc_y_, z = batted_ball_data$"effective_speed", group = -1)) + 
      stat_summary_hex(color="gray", na.rm =TRUE, fun=median) + #<
      scale_fill_distiller(palette = "YlGnBu", direction=1, name = "x") +
      geom_mlb_stadium(stadium_ids = unique(batted_ball_data$team), 
                       stadium_transform_coords = TRUE, stadium_segments = "all", na.rm=TRUE) +
      coord_fixed() +
      theme_void()
      
  
```



```{r}
batted_ball_data %>% mlbam_xy_transformation() %>%
  ggplot(aes(x = hc_x_, y = hc_y_)) +
    geom_hex(color="gray", na.rm=TRUE) +
    scale_fill_distiller(palette = "YlGnBu", direction=1, name = "x") +
    geom_mlb_stadium(stadium_ids = unique(batted_ball_data$team), stadium_transform_coords = TRUE, stadium_segments = "all") +
    coord_fixed() +
    theme_void()
```

```{r}
mean(batted_ball_data$effective_speed)
```


```{r}
test_data <- mlb_batted_balls_2021 %>% filter(home_team == 'SEA')
batted_ball_data <- copy(test_data)
unique(batted_ball_data$game_date)
dates <- length(unique(batted_ball_data$game_date))
```

```{r}
batted_ball_data$team <- rep(c(my_list[[batted_ball_data$home_team[1]]]), each=dim(batted_ball_data)[1])

# batted_ball_data$row_team <- factor(batted_ball_data$row_team, levels = c("COL", "ARI"))

batted_ball_data %>% mlbam_xy_transformation() %>% 
    ggplot(aes(x = hc_x_, y = hc_y_, z = home_runs, group = -1)) + 
      stat_summary_hex(color="gray", na.rm =TRUE, fun=median) + #<
      scale_fill_distiller(palette = "YlGnBu", direction=1) +
      geom_mlb_stadium(stadium_ids = unique(batted_ball_data$team), 
                       stadium_transform_coords = TRUE, stadium_segments = "all", na.rm=TRUE) +
      coord_fixed() +
      theme_void()
```


```{r}
batted_ball_data %>% mlbam_xy_transformation() %>%
  ggplot(aes(x = hc_x_, y = hc_y_)) +
    geom_hex(color="gray", na.rm=TRUE) +
    scale_fill_distiller(palette = "YlGnBu", direction=1) +
    geom_mlb_stadium(stadium_ids = unique(batted_ball_data$team), stadium_transform_coords = TRUE, stadium_segments = "all") +
    coord_fixed() +
    theme_void()
```
```{r}
metric <- c("Launch Angle", 'Hit Distance', "Effective Speed", "Launch Speed")
vals <- c("launch_angle", 'hit_distance_sc', "effective_speed", "launch_speed")
names(vals) <- metric
my_list2 <- as.list(vals)
my_list2
```
```{r}
val <- my_list2[['Launch Speed']]
batted_ball_data$'launch_speed'
```

```{r}
rankings <- c(1:30)
rankings
```

```{r}
ecdf(rankings)(29)

```
```{r}
c("Launch Angle", 'Hit Distance', "Effective Speed", "Launch Speed")
```

```{r}
teams[6]
```
```{r}
mlb_batted_balls_2021 %>% filter(home_team == "WAS")
```
```{r}
teams <- c("PHI", "OAK", "TB", "MIA", "WSH", "SF", "NYY", "SD", "NYM", "SEA", "BOS",
           "HOU", "BAL", "LAA", "TOR", "STL", "CIN", "LAD", "TEX", "DET", "CHC",
           "CWS", "MIL", "CLE", "PIT", "MIN", "KC", "ATL", "ARI", "COL")

# stadiums <- c("Citizens Bank Park", "Oakland Coliseum", "Tropicana Field", "Marlins Park", "Nationals Park", "Oracle Park", 
#              "Yankee Stadium", "Petco Park", "Citi Field", "T-Mobile Park", "Fenway Park", "Minute Maid Park", "Oriole Park at Camden Yards",
#              
#              "Angel Stadium", "Rogers Centre", "Busch Stadium", "Great American Ball Park", "Dodger Stadium", 
#              "Globe Life Field", "Comerica Park", "Wrigley Field", "Guaranteed Rate Field", "Miller Park", "Progressive Field", "PNC Park", 
#              "Target Field", "Kauffman Stadium", "Truist Park", "Chase Field", "Coors Field")

altitude_rankings <- c(1:30)

outfield_walls <- 31 - c(6,25,20,24,13,2,5,7,21,22,1,3,10,16,29,28,14,
                    27,8,15,18,16,23,12,4,11,30,19,9,17)

outfield_walls
```

```{r}
mlb_batted_balls_2021 <- mlb_batted_balls_2021 %>%
  mutate(bases_covered = case_when(
              events == "single" ~ 1,
              events == "double" ~ 2,
              events == "triple" ~ 3,
              events == "home_run" ~ 4,
              TRUE ~ 0)
  )

```

```{r}
mlb_batted_balls_2021 <- mlb_batted_balls_2021 %>%
  mutate(home_runs = case_when(
              events == "home_run" ~ 4,
              TRUE ~ 0)
  )
```

```{r}
mlb_batted_balls_2021 <- mlb_batted_balls_2021 %>%
  mutate(triple = case_when(
              events == "triple" ~ 3,
              TRUE ~ 0)
  )
```

```{r}
mlb_batted_balls_2021 <- mlb_batted_balls_2021 %>%
  mutate(double = case_when(
              events == "double" ~ 2,
              TRUE ~ 0)
  )
```

```{r}
mlb_batted_balls_2021 <- mlb_batted_balls_2021 %>%
  mutate(single = case_when(
              events == "single" ~ 1,
              TRUE ~ 0)
  )
```

```{r}
bases_covered = c()
for (i in 1:30) {
  test <- mlb_batted_balls_2021 %>% filter(home_team == teams[i])
  # data <- sample_n(test, 1000)
  # outliers <- boxplot(data$hit_distance_sc, plot=FALSE)$out
  # data <- data[-which(data$hit_distance_sc %in% outliers),]
  val <- sum(test$bases_covered, na.rm = TRUE)
  bases_covered[i] <- val
}
bases_covered
```
```{r}
home_runs = c()
for (i in 1:30) {
  test <- mlb_batted_balls_2021 %>% filter(home_team == teams[i])
  # data <- sample_n(test, 1000)
  # outliers <- boxplot(data$hit_distance_sc, plot=FALSE)$out
  # data <- data[-which(data$hit_distance_sc %in% outliers),]
  val <- sum(test$home_runs == 4, na.rm = TRUE)
  home_runs[i] <- val
}
home_runs
```
```{r}
triples = c()
for (i in 1:30) {
  test <- mlb_batted_balls_2021 %>% filter(home_team == teams[i])
  # data <- sample_n(test, 1000)
  # outliers <- boxplot(data$hit_distance_sc, plot=FALSE)$out
  # data <- data[-which(data$hit_distance_sc %in% outliers),]
  val <- sum(test$triple == 3, na.rm = TRUE)
  triples[i] <- val
}
triples
```
```{r}
doubles = c()
for (i in 1:30) {
  test <- mlb_batted_balls_2021 %>% filter(home_team == teams[i])
  # data <- sample_n(test, 1000)
  # outliers <- boxplot(data$hit_distance_sc, plot=FALSE)$out
  # data <- data[-which(data$hit_distance_sc %in% outliers),]
  val <- sum(test$double == 2, na.rm = TRUE)
  doubles[i] <- val
}
doubles
```

```{r}
singles = c()
for (i in 1:30) {
  test <- mlb_batted_balls_2021 %>% filter(home_team == teams[i])
  # data <- sample_n(test, 1000)
  # outliers <- boxplot(data$hit_distance_sc, plot=FALSE)$out
  # data <- data[-which(data$hit_distance_sc %in% outliers),]
  val <- sum(test$single == 1, na.rm = TRUE)
  singles[i] <- val
}
singles
```


```{r}
hit_distance = c()
for (i in 1:30) {
  test <- mlb_batted_balls_2021 %>% filter(home_team == teams[i])
  # data <- sample_n(test, 1000)
  # outliers <- boxplot(data$hit_distance_sc, plot=FALSE)$out
  # data <- data[-which(data$hit_distance_sc %in% outliers),]
  val <- mean(test$hit_distance_sc, na.rm = TRUE)
  hit_distance[i] <- val
}
hit_distance
```

```{r}
effective_speed = c()
for (i in 1:30) {
  test <- mlb_batted_balls_2021 %>% filter(home_team == teams[i])
  # data <- sample_n(test, 1000)
  # outliers <- boxplot(data$effective_speed, plot=FALSE)$out
  # data <- data[-which(data$effective_speed %in% outliers),]
  val <- mean(test$effective_speed, na.rm = TRUE)
  effective_speed[i] <- val
}
effective_speed
```

```{r}
df <- data.frame(teams, altitude_rankings, outfield_walls, hit_distance, effective_speed, bases_covered, 
                 home_runs, triples, doubles, singles)
names(df) <- c("team", "altitude_rankings", "outfield_walls", "hit_distance", "effective_speed", "bases_covered", 
               "home_runs", "triples", "doubles", "singles")
# ecdf(df$effective_speed)(89.44370)
```

```{r}
#install.packages('toOrdinal')
library(toOrdinal)
toOrdinal(5)
```

```{r}
toOrdinal(ecdf(df$effective_speed)(df$effective_speed[df$team == 'PHI']) * 100)
```
```{r}
df
```

```{r}
saveRDS(df,file="percentiles_df.rds")
```

```{r}
saveRDS(mlb_batted_balls_2021,file = "mlb_data.rds")
```

```{r}
#install.packages('floor')
floor(66.6666)
```

